---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Salaries range widely, from $0 up to $197,200, with lots of clusters of the same salaries in groups. Such as $65,500, $74,500, $83,500, $101,500, $121,500, etc, suggesting that the White House uses standardized pay for many staff positions compared to making salaries individually. From a reporting standpoint the $0.00 salaries are the most notable. Several senior officials including Marco Rubio the National Security Advisor are listed with $0 salaries. Some of these officials are paid through a different part of the Federal Government, Rubio's actual salary comes from his role as Secretary of State also around $260,00 a year. Employees listed under Detailees break the pattern of clustered salaries, they are paid in non-rounded numbers such as $105,829, which could mean they are being paid from a separate organization as well. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean (SALARY),
            media_salary = median (SALARY)) 
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean(SALARY),
            median_salary = median(SALARY),
            min_salary = min(SALARY),
            max_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary is higher than the median salary, which could mean that the data is skewed slightly by some employee's higher earnings pulling up the average. The maximum salary shows the senior staff at the top of the White House pay scale, like assistant to the president. The minimum salary being $0 is not actually correct, Cabinet officials like Marco Rubio whose salaries are paid separately could potentially pull the mean and median salaries upward or downward. But those salaries are excluded from the White House Budget. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

A story idea I had is: Why do some powerful people in the White House show a $0 dollar salary and how does this affect where political power lies in the house? Without doing research some people might think they aren't getting paid. Officials like Marco Rubio hold a lot of influence inside the White House although being paid by a separate agency. Which brings the questions: Does being paid elsewhere affect oversight? How many roles is one person allowed to hold at once? To take the next steps I would identify all $0 salary employees and look up what agency is compensating them, check whether it's a legal or historical reason some officials have more than one title through one paying agency. I would then reach out to a White House or governance expert, maybe a political science professor. 
