---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- 
read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> 
  mutate(
    violent_rate_2019 = violent2019 / pop2019 * 10000,
    property_rate_2019 = property2019 / pop2019 * 10000,
    violent_rate_2020 = violent2020 / pop2020 * 10000,
    property_rate_2020 = property2020 / pop2020 * 10000
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

In 2019 Baltimore led the violent crime rate (183 per 10,000), but Ocean city was the highest in 2020, jumping to 240 per 10,000 and Baltimore's rate that year dropped to 160. Small towns like Williamsport and Kitzmiller also show up near the top violent crime rate, although they have tiny populations. For property crime rate, Ocean City has the highest rate in both years by length, over 1,000 per 10,000 in 2019 and even more in 2020. Baltimore's property rate also fell between these years. This doesn't fully add up to me, I believe that smaller populations are inflating these rates. The town of Kitzmiller population of 234, and other tiny towns are being recorded in the top 10 highest rates for just a handful of incidents based on their small population. Compared to Baltimore, which definitely has the highest raw crime counts, saw a decline in both categories from 2019-2020. 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> 
  mutate(
    violent_pct_change = (violent_rate_2020 - violent_rate_2019) / violent_rate_2019 * 100,
    property_pct_change = (property_rate_2020 - property_rate_2019) / property_rate_2019 * 100
  )
    
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

Something interesting I noticed is that there are many outliers in the data that could be throwing off perceptions. Most of the biggest percentage changes, increases and decreases, come from overall small towns where just one or two incidents can look like a 100% or even an almost 900% change, which draws attention but is actually not that meaningful. I believe you could make a real story out of Baltimore where both violent and property crime rates dropped a lot in such a large population, this could just be from covid though. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Violent and property crime rates in Baltimore fell in 2020, with violent crime rates dropping by almost 13% and property crime rates dropping by nearly 26% from the previous year. 

The decline in Maryland's largest city stands out in a statewide dataset comparing all of Maryland's cities from 2019 to 2020. Baltimore's rate had a higher percentage decrease than most other largely populated cities. Some small towns with populations in the hundreds showed very dramatic percentage changes due to a handful of incidents and such a little population. Baltimore's numbers are based on a population of nearly 600,000, making the decrease more likely to reflect a pattern. The timing lines up with the 2020 Covid-19 pandemic, where lock downs and reduced public activity limited opportunities for both types of crimes. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: 
```{r}
 md_cities_percentiles <- md_cities_rates |> 
  mutate(percentile = ntile(violent_rate_2020, 100)) 
```
- There are 10 cities above the 90th percentile 

```{r}
md_cities_percentiles |> summarize(iqr_violent_2020 = IQR(violent_rate_2020)) 

```

Do the results change your best story idea from the previous assignment? Why or why not?

The results showing the IQR and percentile do not change my story idea. Baltimore ranks in the top 10 cities statewide for violent crime rate, which is consistent with the idea in my original lede and nutgraf about Baltimore's crime trends being significant given it's size and crime levels relative to Maryland cities. The IQR range is 42 and Baltimore falls much higher, which supports that Baltimore's crime decline is notable in the context of a larger, high-crime city. 

