---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse) 
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset

```{r}
sample_5 <- mandms |> 
  sample_n(5)

sample_5
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the red or black number of M&Ms

```{r}
sample_5 |> 
  summarize(
    mean_red = mean(red),
    sd_red = sd(red)
  )
```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
mandms |> 
  summarize(
    mean_red = mean(red),
    sd_red = sd(red)
  )
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

They are somewhat similar but not close. My sample of 5 had a mean of 62.8 red M&Ms and a standard deviation of 10.71, compared to the population mean of 72.98 and standard deviation of 16.46. My sample mean is lower, and my standard deviation is about 6 lower, because my 5 rows happened to be mostly low values and didn't capture the full population. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms data set and about the act of random sampling?

My random sample underestimated both the average number of M&Ms and how much the counts vary. The full dataset ranges from 45 5o 117 red M&Ms, but my sample only covered 45 to 73, so it missed the entire upper half of the data. This shows that a small random sample can give a misleading picture of the population just by chance, even when sampling fairly.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
