---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- mean(ces$evangelical)

population_proportion
```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

The population proportion of white Evangelicals is about 0.105, or 10.5% of respondents. Which is much smaller than the unaffiliated proportion of about 0.36, unaffiliated americans are roughly three and a half times as common in the data. Because White Evangelicals are a much smaller group, small samples might include only a few of them, so their results could be pretty farr from the real 10%. 

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)
sample_50 <- ces |> sample_n(50)
mean(sample_50$evangelical) 
```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

My sample got 16% white evangelical but the real number is about 10.5% so it was off by about 5.5 points. That's a bigger miss than it sounds, because the group is so small. Being off by the same mount for the unaffiliated group in class wouldn't matter as much, since that group is 36% of the population. 

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966)

sample_500 <- ces |> 
  sample_n(500)

sample_500 |> 
  summarize(prop = mean(evangelical))
```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

My sample of 500 got 9% white evangelical which is only about 1.5 points off from the real 10.5%. That's much closer than my sample of 50, which was off by 5.5 points. A bigger sample gave a more accurate estimate. 

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share White Evangelical",
    subtitle = "Red line = true population proportion (about 10.5%). Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

The n=50 histogram has most polls landing near 10%, but it's a little lopsided. On the left it stops at 0%, but on the right some polls go past 20%. The unaffiliated histogram from class was more even, phasing off about the same on both sides of 36%. The difference is that a sample of 50 only has about 5 white evangelicals on average. a poll cant find fewer than 0 but it could find 10 or more by chance so the results only stretch out in one direction. 

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

With 500 people, the histogram got much narrower. Polls went from a range of about0% to 24% down to about 7% to 15%, much closer to the real 10.5% and the shape became more even. The unaffiliated group in class narrowed by more points, from about 20%-55% down to 31%-42%, because it started out more spread out. But the improvement matters more for white evangelicals because missing by a few points is a big deal when the real number is only 10.5%. 

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

A poll of 200 people would include only about 21 white evangelicals, since they're about 10.5% of the population. That's a very small group to base a story on, so just a few people answering differently could change the results a lot. this shows that results for small groups within a poll are much less reliable, and journalists should be careful about making big claims based on them. 

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

I would want to know how many white evangelicals were actually in the poll. outof 400 people, there would probably only be about 42, so the 78% is based on a small group and could be pretty far off. Since a campaign shared it, i'd want to know who ran the poll and how they picked people. I would also check whether other polls showed something similar. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
